---
title: "Lab 4 Normal Distribution"
author: "your name"
output:
  html_document:
    css: lab.css
    highlight: pygments
    theme: cerulean
---

```{r echo=F, message=F}
library(tidyverse)
library(openintro)
library(plotly)
```

```{r, message=F, echo=FALSE}
nc <- read.csv("/cloud/project/nc.csv")

nc_ft <- nc %>% 
  filter(premie=="full term")
```

1.  Describe the distribution (shape, center, variation)

```{r, warning=FALSE,echo=F }
ggplot(nc_ft, aes(weight ))+
  geom_histogram(binwidth=0.25,color="white", fill="steelblue")+
  scale_x_continuous(limits=c(3, 12), breaks=c(seq(3,12, by=1)))
```

2.  Based on the this plot, does it appear that the data follow a nearly normal distribution?

```{r, echo=FALSE, warning=FALSE}
mean_nc = mean(nc_ft$weight)
sd_nc = sd(nc_ft$weight)

ggplot(nc_ft, aes(x=weight)) +
  geom_blank() +
  geom_histogram(binwidth=0.25,color="white", fill="lightblue", aes(y=..density..)) +
  stat_function(fun = dnorm, args = c(mean = mean_nc, sd = sd_nc), col = "tomato")+
  scale_x_continuous(limits=c(3, 12), breaks=c(seq(3,12, by=1)))+
  geom_vline(xintercept=c(mean_nc,mean_nc+sd_nc, mean_nc+2*sd_nc, mean_nc+3*sd_nc, mean_nc-sd_nc, mean_nc-2*sd_nc, mean_nc-3*sd_nc), linetype="dashed")
```


3.  Below is a comparison of the simulated data (in red) along with the `nc_ft` real data. Do all of the points fall on the line?  How does this plot compare to the probability plot for the real data? 

```{r, echo=F}
sim_norm <- data.frame(sim_values = rnorm(n = nrow(nc_ft), mean = mean_nc, sd = sd_nc))

ggplot(data = nc_ft, aes(sample = weight)) + 
  geom_point(stat = "qq", color = "steelblue", size=0.5)+
  geom_line(data=sim_norm,aes(sample=sim_values),stat = "qq", color="tomato")
```

4.  Does the normal probability plot for the `weight`  look similar to the 8 plots 
    created for the simulated data?  That is, do the plots provide evidence that the
    `weight` is nearly normal?
    
```{r qqnormsim, echo=F}
source("http://www.jkarreth.net/files/qqnormsim.R")  ## loading the qqnormsim function
qqnormsim(nc_ft$weight)
```

5.  Use both methods above to determine the probability that for a birth `weight` less than 7 pounds. Comment on why they are differeny (Hint: look at your graphics)

**More Practice**

We have just created a procedure to assess a quantitative variable if follows a Normal Distribution.

**Step 1**: Graph a density histogram and a normal curve (with mean and sd generated from data set)
```{r, echo=F, warning=F}
mean_nc = mean(nc_ft$weight)
sd_nc = sd(nc_ft$weight)
ggplot(nc_ft, aes(x=weight)) +
  geom_blank() +
  geom_histogram(binwidth=0.25,color="white", fill="lightblue", aes(y=..density..)) +
  stat_function(fun = dnorm, args = c(mean = mean_nc, sd = sd_nc), col = "tomato")+
  scale_x_continuous(limits=c(3, 12), breaks=c(seq(3,12, by=1)))+
  geom_vline(xintercept=c(mean_nc,mean_nc+sd_nc, mean_nc+2*sd_nc, mean_nc+3*sd_nc, mean_nc-sd_nc, mean_nc-2*sd_nc, mean_nc-3*sd_nc), linetype="dashed")
```

**Step 2**:  Graph a QQ Plot of the actual data and simulated normal data
```{r, echo=F}
sim_norm <- data.frame(sim_values = rnorm(n = nrow(nc_ft), mean = mean_nc, sd = sd_nc))
ggplot(data = nc_ft, aes(sample = weight)) + 
  geom_point(stat = "qq", color = "steelblue", size=0.5)+
  geom_line(data=sim_norm,aes(sample=sim_values),stat = "qq", color="tomato")
```

**Decision**: If normal curve and simulated normal data are similar, the data can be approximated by a normal distribution.

Analysize `mage` and `gained` using the 2 steps above and decide if they can be modeled with the Normal Distribution.  

6.  Complete the analysis outlined above for `mage` and make a decision if `mage` can be modeled with the Normal Distribution (i.e., Step 1 and Step 2).  Test your decision by calculating the probability that a mother's age is less than 21.

7.  Complete the analysis outlined above for `gained` and make a decision if `gained` can be modeled with the Normal Distribution.  Test your decision by calculating the probability that a mother's weight gained is more than 40.



